```{r fig-consort, fig.width = 7.086614166, fig.height = 4.330708657, fig.cap = 'Flow diagram of the analysis inclusion process for the CovILD and HACT studies.'}

figures$consort$plot

```

__Figure \@ref(fig:fig-consort). Flow diagram of the analysis inclusion process for the CovILD and HACT studies.__



```{r fig-sympt-recovery, fig.width = 7.086614166, fig.height = 8.661417314, fig.cap = 'Symptom-specific recovery times in ambulatory COVID-19.'}

figures$sympt_recovery$plot

```

__Figure \@ref(fig:fig-sympt-recovery). Symptom-specific recovery times in ambulatory COVID-19.__



```{r fig-sympt-mds, fig.width = 7.086614166, fig.height = 7.086614166, fig.cap = 'Self-reported smell and taste disorders are isolated symptoms of long COVID and PASC.'}

figures$sympt_mds$plot

```

__Figure \@ref(fig:fig-sympt-mds). Self-reported smell and taste disorders are isolated symptoms of long COVID and PASC.__



```{r fig-apriori, fig.width = 7.086614166, fig.height = 7.480314953, fig.cap = 'High extent of co-occurrence of smell and taste disorders in long COVID and PASC.'}

figures$apriori$plot

```

__Figure \@ref(fig:fig-apriori). High extent of co-occurrence of smell and taste disorders in long COVID and PASC.__



```{r fig-clustering, fig.width = 7.086614166, fig.height = 7.086614166, fig.cap = 'Clustering of ambulatory COVID-19 individuals by symptom-specific recovery times.'}

figures$clustering$plot

```

__Figure \@ref(fig:fig-clustering). Clustering of ambulatory COVID-19 individuals by symptom-specific recovery times.__



```{r fig-sympt-clusters, fig.width = 7.086614166, fig.height = 7.086614166, fig.cap = 'Duration of neurocognitive and respiratory symptoms, fatigue, smell and taste disorders differs between the COVID-19 recovery clusters.'}

figures$sympt_clusters$plot

```

__Figure \@ref(fig:fig-sympt-clusters). Duration of neurocognitive and respiratory symptoms, fatigue, smell and taste disorders differs between the COVID-19 recovery clusters.__



```{r fig-base-clusters, fig.width = 7.086614166, fig.height = 8.267716527, fig.cap = 'COVID-19 recovery clusters differ in sex distribution, comorbidity and daily medication rates.'}

figures$base_clusters$plot

```

__Figure \@ref(fig:fig-base-clusters). COVID-19 recovery clusters differ in sex distribution, comorbidity and daily medication rates.__



```{r fig-recovery-clusters, fig.width = 7.086614166, fig.height = 4.724409444, fig.cap = 'Good clinical, mental and psychosocial recovery in individuals with persistent smell and taste disorders'}

figures$recovery_clusters$plot

```

__Figure \@ref(fig:fig-recovery-clusters). Good clinical, mental and psychosocial recovery in individuals with persistent smell and taste disorders__

